### 2017-01-19 manuscript.rmd
updated hard citations to bib

@@

How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo). To address this topic, Levin *et al.* <sup>1</sup> analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions. First, animal development is characterized by a well-defined mid-developmental transition. Second, this transition helps explain the evolution of features observed among distantly related animals. Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development <sup>2,3</sup>. Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a long-sought operational definition of phyla. We previously described some concerns with their interpretations of these results <sup>4</sup>. Here we directly address problems with the analyses behind their evolutionary conclusions. Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* <sup>1</sup> indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly). Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology <sup>5,6</sup>.

To understand our concerns with their analyses, it is helpful to first outline Levin *et al.*'s published methods and results. Although there are mature methods and tools for statistical analysis of character evolution <sup>7</sup> and gene expression <sup>8</sup>, Levin *et al.* <sup>1</sup> drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development while the goodness of fit to these patterns was not considered, and alternative patterns were not evaluated. For each species pair, they then identified the orthologs shared by these species (shared orthologs vary from pair to pair). They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

@@

Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental <sup>9</sup> and functional genomic traits <sup>10</sup>. This is particularly true for evolutionary analyses of quantitative gene expression <sup>11</sup>. Pairwise comparisons result in the same evolutionary changes being counted multiple times in each species pair, as for the ctenophore comparisons that were mistaken for a global pattern here. This is a well understood property of pairwise comparisons that has been specifically addressed by phylogenetic comparative methods <sup>12</sup>. This particular case illustrates the shortcomings of *ad hoc* rank-based methods <sup>3</sup> and pairwise comparisons <sup>1</sup> for the study of the evolution of gene expression. Phylogenetic analyses of character evolution provide not only a way to avoid these problems, but a richer context for interpreting the biological implications of the results.

@@

>>>>

# Comparing gene expression across species

Dunn, Zapata, Munro, Siebert, Hejnol

@@

How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo). To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions. First, animal development is characterized by a well-defined mid-developmental transition. Second, this transition helps explain the evolution of features observed among distantly related animals. Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; DomazetLoso:2010cn]. Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a long-sought operational definition of phyla. We previously described some concerns with their interpretations of these results [@Hejnol:2016hw]. Here we directly address problems with the analyses behind their evolutionary conclusions. Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly). Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology [@Martindale:2015ve, Dunn:2015ij].

To understand our concerns with their analyses, it is helpful to first outline Levin *et al.*'s published methods and results. Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development while the goodness of fit to these patterns was not considered, and alternative patterns were not evaluated. For each species pair, they then identified the orthologs shared by these species (shared orthologs vary from pair to pair). They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

@@

Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental [@Telford:2003tsa] and functional genomic traits [@Hejnol:2015gr]. This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa]. Pairwise comparisons result in the same evolutionary changes being counted multiple times in each species pair, as for the ctenophore comparisons that were mistaken for a global pattern here. This is a well understood property of pairwise comparisons that has been specifically addressed by phylogenetic comparative methods [@Felsenstein:1985ua]. This particular case illustrates the shortcomings of *ad hoc* rank-based methods [@DomazetLoso:2010cn] and pairwise comparisons [@Levin:2016bp] for the study of the evolution of gene expression. Phylogenetic analyses of character evolution provide not only a way to avoid these problems, but a richer context for interpreting the biological implications of the results.

@@

Thanks to
